---
title: "Debate in the Wild: Data Preparation, Data Summaries, and Targeted Model"
output: html_document
  # html_document:
  #   fig_caption: yes
  #   number_sections: yes
  #   toc: yes
---

This R markdown provides the basis for our manuscript exploring the effects of linguistic behaviors and personal traits on persuasiveness using <a href = "http://www.iq2us.org" target="_blank">the Intelligence Squared U.S.</a> corpus, first introduced by Paxton and Dale (2014, *Proceedings of the Cognitive Science Society*). This is the first of two R markdowns that accompanies this manuscript.

To run completely from scratch, you will need the following files:

* `./data/all-debate-data.csv`
* `./supplementary/libraries-and-functions_debate-in-the-wild.R`
* `./supplementary/missing-data_debate-in-the-wild.R`

Additional files required for the data analysis will be generated at various steps throughout the code.

**NOTE**: The chunks of code in many sections need only be run once, since the manipulated data will be saved to new files. As a result, these chunks are currently set to `eval = FALSE`. Bear this in mind if these data need to be re-calculated.

**Code written by**: A. Paxton (University of California, Berkeley)
<br>**Date last modified**: 23 June 2016

*** 

# Global preliminaries

This section takes care of data import and preparation. As a new analysis of the IQ2 corpus (Paxton & Dale, 2014, *Proceedings of the Cognitive Science Society*), the original data file includes coded variables from prior work.

***

## Load libraries and functions

This section sets the working directory, loads in the requisite libraries, and creates new functions for the analyses.

```{r global-prelims, warning = FALSE, error = FALSE, message = FALSE}

# clear up our workspace
rm(list=ls())

# set working directory for the first time
setwd('/debate-in-the-wild/')

# load in functions and libraries
source('./supplementary/libraries-and-functions_debate-in-the-wild.R')

```

***

# Data preparation

While we have the complete dataset including all debates, we still need to prepare the data a bit before we can analyze everything. This section takes the master dataframe and gets it ready for data analysis.

***

## Preparing debater-only subset

After having cleaned up the entire debate dataset, we here prepare the subset of data in which we're interested: turns at which only the debaters are speaking.

***

### Preliminaries

```{r prep-debater-prelims, eval = FALSE, warning = FALSE, error = FALSE, message = FALSE}

# clean everything up
rm(list = ls())

# read in our source files
setwd('/debate-in-the-wild/')
source('./supplementary/libraries-and-functions_debate-in-the-wild.R')

# read in the concatenated dataset
debateList = read.csv('./data/all-debate-data.csv', header = TRUE, sep = ",", 
                      row.names = NULL, fill = FALSE)

```

***

### Winnow data and update NAs

This section removes turns by the moderator (`group` = `M`) and other speakers (`group` = `O`), then transforms missing data to `NA`.

```{r winnow-and-NAs, eval = FALSE}

# keep only data where debaters are talking
debaters = debateList[debateList$group!='O',]
debaters = debaters[debaters$group!='M',]

# convert 909 (for photo ratings) and 999 (for in-depth voting details) to NA
debaters[debater.ratings][debaters[debater.ratings]==909] <- NA
debaters[detailed.variables][debaters[detailed.variables]==999] <- NA

```

***

### Recode groups and clean up

Let's create numeric codes for our debater groups ("against" = `1`, "for" = `2`), remove variables that were unavailable for all debates, and drop lines with missing data.

```{r recode-debater-groups, eval = FALSE}

# recode debater groups
debaters$group = as.numeric(debaters$group)

# remove the in-depth debate metrics
debaters = debaters[names(debaters) %notin% detailed.variables]

# drop any lines that have missing data
debaters = debaters[complete.cases(debaters),]

```

***

### Aggregate data at speaker level

Since the data were created in the B(eo)W(u)LF format (*by-word long-form*; Paxton & Dale, 2013, *arXiv*), the data are separated at the word level. We here aggregate data at the speaker level and export the data.

```{r aggregate-data, eval = FALSE}

# identify the LIWC variables
all.vars = names(debaters)
liwc.variables = all.vars[all.vars %notin% c(debater.ratings,
                                             desc.variables,
                                             text.variables,
                                             outcome.variables,
                                             summedLIWC.variables)]

# grab the outcome variables and debater ratings
outcome.data = debaters %>% 
  select(one_of(c(desc.variables,outcome.variables, debater.ratings))) %>%
  group_by(debate, group, speaker) %>%
  summarise_each( funs( mean ))

# create turn-level transcript
reconstituted.transcript = debaters %>% 
  group_by(debate, group, speaker) %>% 
  summarise(transcript = toString(word, sep=" ")) 

# aggregate base-100 LIWC variables
aggregated.liwc = debaters %>% 
  select(one_of(c(liwc.variables,desc.variables))) %>%
  group_by(debate, group, speaker) %>% 
  summarise_each( funs( sum(as.numeric(.))/100 ))

# sum the single-count variables
summed.liwc = debaters %>% 
  select(one_of(c(summedLIWC.variables,desc.variables))) %>%
  group_by(debate, group, speaker) %>% 
  summarise_each( funs( sum( as.numeric(.) )))

# bring together the individual dataframes
debaters = join(outcome.data, reconstituted.transcript,
                by = c('debate','group','speaker'))
debaters = join(debaters, summed.liwc, 
                by = c('debate','group','speaker'))
debaters = join(debaters, aggregated.liwc,
                by = c('debate','group','speaker'))

```

***

### Export aggregated speaker-level dataset

Let's save the speaker-level data to a CSV.

```{r export-turn-level-data, eval = FALSE}

# export the speaker-level data
write.csv(debaters, './data/speaker_level-debate-data.csv', row.names=FALSE)

```

***

## Standardize variables

In order to interpret the estimates of our forthcoming mixed-effects models as effect sizes (cf. Keith, 2008), we need to center and scale them before entering them into our model.

***

### Preliminaries

Let's clean up the workspace, read in the source file, and import the data.

```{r standardize-prelim, eval = FALSE, warning = FALSE, error = FALSE, message = FALSE}

# clean everything up
rm(list = ls())

# read in our source files
setwd('/debate-in-the-wild/')
source('./supplementary/libraries-and-functions_debate-in-the-wild.R')

# read in the concatenated dataset
debaters = read.csv('./data/speaker_level-debate-data.csv', header = TRUE, sep = ",", 
                    row.names = NULL, fill = FALSE)

```

***

### Create proportion terms

Let's create new variables that treat our predictors as proportions of total words spoken by the speaker (`WC`). We also include the plural first-person pronoun category (`we`).

```{r create-proportion-terms, eval = FALSE}

# identify the LIWC variables
liwc.variables = names(debaters)[names(debaters) %notin% c(debater.ratings,
                                             desc.variables,
                                             text.variables,
                                             summedLIWC.variables,
                                             outcome.variables)]

# calculate variables as proportion of total words spoken by the speaker
proportion.vars = debaters %>%
  ungroup() %>%
  select(one_of(desc.variables,liwc.variables), WC) %>%
  mutate_each(funs( as.numeric(. / WC) ), -one_of(desc.variables, WC)) %>%
  select(-WC) %>%
  setNames(paste0('prop.', names(.)))
proportion.vars = plyr::rename(proportion.vars, 
                               c("prop.debate" = "debate",
                                 "prop.group" = "group",
                                 "prop.speaker" = "speaker"))

# merge with original dataset
debaters = join(debaters,proportion.vars,by=c(desc.variables))

```

***

### Create interaction terms

We here create standardized interaction terms that we can use for during the analysis stage in order to properly create models for both standardized and unstandardized models.

```{r create-group-interaction-terms, eval = FALSE}

# create interactions with group membership
group.interaction.vars = debaters %>%
  ungroup() %>%
  select(one_of(desc.variables,
                names(proportion.vars),
                liwc.variables,
                debater.ratings)) %>%
  mutate_each(funs( as.numeric(. * group) ), -one_of(desc.variables)) %>%
  setNames(paste0('group.', names(.)))
group.interaction.vars = plyr::rename(group.interaction.vars, 
                                      c("group.debate" = "debate",
                                        "group.group" = "group", 
                                        "group.speaker" = "speaker"))

# merge with existing dataset
debaters = join(debaters,group.interaction.vars,by=c(desc.variables))

```

***

### Standardize dataset

This step will create a centered and standardized dataset for our standardized models.

```{r standardize-main-effects, eval = FALSE}

# scale variables
debaters.st = debaters %>% ungroup() %>%
  mutate(speaker = as.integer(speaker)) %>%
  purrr::keep(is.numeric) %>%
  mutate_each( funs( as.numeric( scale(.) )))

```

***

### Save plotting and standardized datasets

We save separate versions of the new dataset, unstandardized (`./data/plotting-debate-data.csv`) and standardized (`./data/standardized-debate-data.csv`).

```{r save-plot-and-st-data, eval = FALSE}

write.csv(debaters, './data/plotting-debate-data.csv', row.names=FALSE)
write.csv(debaters.st, './data/standardized-debate-data.csv', row.names=FALSE)

```

***

# Data analysis

***

## Preliminaries

Let's clean everything up again and get ready for data analysis

```{r analysis-prelim, warning = FALSE, error = FALSE, message = FALSE}

# clean everything up
rm(list = ls())

# read in our source files
setwd('/debate-in-the-wild/')
source('./supplementary/libraries-and-functions_debate-in-the-wild.R')

# read in the plotting and standardized datasets
debaters.plotting = read.csv('./data/plotting-debate-data.csv', header = TRUE, sep = ",", 
                             row.names = NULL, fill = FALSE)
debaters.st = read.csv('./data/standardized-debate-data.csv', header = TRUE, sep = ",", 
                       row.names = NULL, fill = FALSE)

```

***

## Some summary statistics

Let's take a peek at some of the attributes of the corpus before jumping in.

***

### How many debates do we have?

```{r num-debates}
# how many debates do we have?
total.debates = length(unique(debaters.plotting$debate))
total.debates
```

***

### How many speakers are included?

```{r num-speakers}
# how many unique speakers do we have?
total.speakers = length(unique(debaters.plotting$speaker))
total.speakers
```

***

### How many words are included?

```{r num-words}
# how many words do we have?
total.words = sum(debaters.plotting$WC)
print(total.words)
```

***

### How often does each team win?

```{r proportion-of-wins}

# get one line per debate
agg.data = debaters.plotting %>%
  select(debate,winner,deltaV) %>%
  group_by(debate,winner,deltaV) %>%
  summarise_each(funs(mean))

# what proportion of times does the "for" group win?
for.wins = sum(agg.data$winner)/length(agg.data$winner)
for.wins
```

***

### What's the range of net gains for the "for" group relative to the "against" group?

```{r range-deltaV}

# what's the range of deltaV?
range.deltaV = range(agg.data$deltaV)
range.deltaV

```

```{r plot-hist-deltaV, eval=TRUE, echo=FALSE, fig.cap="Debate outcomes (∆V) as a histogram and density function. Raw counts of each bin are represented as a gradient color scale from light blue (more occurrences) to dark blue (fewer occurrences).",fig.width=4, fig.height=3, fig.align='center'}

ggplot(agg.data,aes(x = deltaV)) +
  geom_histogram(aes(y = ..density..,
                     fill = ..count..),
                 binwidth = 5) + 
  geom_density() +
  labs(title = "Debate Outcomes",
       y = "Density",
       x = "∆V") +
  scale_fill_gradient("Count") +
  theme(legend.position = "bottom")

ggsave('./figures/debate_outcome_hist.png')

```
**Figure**: Debate outcomes (∆V) as a histogram and density function. Raw counts of each bin are represented as a gradient color scale from light blue (more occurrences) to dark blue (fewer occurrences).

***

### What's the mean net gain for the "for" group relative to the "against" group?

```{r mean-deltaV}

# what's the mean net gain for the "for" group (deltaV)?
mean.deltaV = mean(agg.data$deltaV)
mean.deltaV

```

***

## Differentiating between winning strategies

```{r winning-strategies-lmer-model}

# raw (unstandardized) model
deltaV.raw.lm = lmer(deltaV ~ group + prop.tentat + prop.we +
                           attractive + prop.negemo + 
                           group.prop.tentat + group.prop.we +
                           group.attractive + group.prop.negemo +
                           (1 + group + prop.tentat + attractive | winner),
                          data = debaters.plotting, REML = FALSE)

# standardized model
deltaV.st.lm = lmer(deltaV ~ group + prop.tentat + prop.we +
                           attractive + prop.negemo + 
                           group.prop.tentat + group.prop.we +
                           group.attractive + group.prop.negemo +
                           (1 + group + prop.tentat + attractive | winner),
                         data = debaters.st, REML = FALSE)

# raw (unstandardized) model results
pander_lme(deltaV.raw.lm, stats.caption = TRUE)

# standardized model results
pander_lme(deltaV.st.lm, stats.caption = TRUE)

```

```{r plot-group-X-attractive-interaction, eval=TRUE, echo=FALSE, fig.cap="Group membership, attractiveness, and winning side.",fig.width=4, fig.height=3, fig.align='center'}

# plot the attractiveness interaction
plot.attr.int = ggplot(data=debaters.plotting, 
                       aes(y = deltaV, 
                           x = attractive, 
                           color = as.factor(group))) +
  geom_smooth() +
  theme(legend.position = "none") +
  scale_color_manual(name="Debater Group",
                     breaks=c("1","2"),
                     values = c('red3','green4'),
                     labels=c("Against", "For")) +
  ylim(c(-80,80)) +
  ylab("∆V") + xlab("Attractiveness")

```
 
```{r plot-group-X-negemo-interaction, eval=TRUE, echo=FALSE, fig.cap="Group membership, negativity, and net change in votes.",fig.width=4, fig.height=3, fig.align='center'}

# plot the negativity interaction
plot.neg.int = ggplot(data=debaters.plotting, 
                      aes(y = deltaV, 
                          x = prop.negemo, 
                          color = as.factor(group))) +
  geom_smooth() +
  #labs(title="Negativity") +
  theme(legend.position = "none") +
  scale_color_manual(name="Debater Group",
                       breaks=c("1","2"),
                       values = c('red3','green4'),
                       labels=c("Against", "For")) +
  ylim(c(-80,80)) +
  ylab(" ") + xlab("Negativity")

```

```{r plot-group-X-hedge-interaction, eval=TRUE, echo=FALSE, fig.cap="Group membership, hedging, and net change in votes.",fig.width=4, fig.height=3, fig.align='center'}

# plot the hedging interaction
plot.hedge.int = ggplot(data=debaters.plotting, 
                        aes(y = deltaV, 
                            x = prop.tentat, 
                            color = as.factor(group))) +
  geom_smooth() +
  #labs(title="Hedging") +
  theme(legend.position = "none") +
  scale_color_manual(name="Debater Group",
                     breaks=c("1","2"),
                     values = c('red3','green4'),
                     labels=c("Against", "For")) +
  ylim(c(-80,80)) +
  ylab("∆V") + xlab("Hedging")

```

```{r plot-group-X-we-interaction, eval=TRUE, echo=FALSE, fig.cap="Group membership, first-person plural pronouns, and net change in votes.",fig.width=4, fig.height=3, fig.align='center'}

# plot the "we" interaction
plot.we.int = ggplot(data=debaters.plotting, 
                     aes(y = deltaV, 
                         x = prop.we, 
                         color = as.factor(group))) +
  geom_smooth() +
  #labs(title="First-Person Plural Pronouns") +
  theme(legend.position = "none") +
  scale_color_manual(name="Debater Group",
                     breaks=c("1","2"),
                     values = c('red3','green4'),
                     labels=c("Against", "For")) +
  ylim(c(-80,80)) +
  ylab(" ") + xlab("First-Person Plural Pronouns")

```

```{r plot-all-interactions, fig.cap="Interactions for the effects of each of the four targeted effects (attractiveness, hedging, negativity, and first-person plural pronouns) and group membership (green = 'for', red = 'against') on outcome (∆V). For hedging, negativity, and first-person plural pronouns, these values are calculated as the proportion of all utterances made up by each category. Again, ∆V is calculated as the pre- to post-debate gains in votes made by the 'for' group relative to the 'against' group. A positive ∆V means that the 'for' group gained more votes than the 'against' group, while a negative ∆V indicates that the 'against' group won more votes relative to the 'against' group.",fig.width=6, fig.height=6, fig.align='center', eval = TRUE, echo = FALSE}

# create a master legend
master.legend = gtable_filter(ggplot_gtable(
  ggplot_build(plot.we.int + theme(legend.position="bottom"))), 
  "guide-box")

# arrange the plots
grid.arrange(
  top=textGrob("Outcomes by Targeted Effects\nand Group Membership",
               gp=gpar(fontsize=14)),
  plot.attr.int,
  plot.neg.int,
  plot.hedge.int,
  plot.we.int,
  ncol = 2
)

# save to file
ggsave('./figures/interactions-for-targeted-effects-on-deltaV.png',
       units = "in", width = 6, height = 6,
       grid.arrange(
         top=textGrob("Outcomes by Targeted Effects\nand Group Membership",
                      gp=gpar(fontsize=14)),
         plot.attr.int,
         plot.neg.int,
         plot.hedge.int,
         plot.we.int,
         ncol = 2
       ))

```

**Figure**: Interactions for the effects of each of the four targeted effects (attractiveness, hedging, negativity, and first-person plural pronouns) and group membership (green = 'for', red = 'against') on outcome (∆V). For hedging, negativity, and first-person plural pronouns, these values are calculated as the proportion of all utterances made up by each category. Again, ∆V is calculated as the pre- to post-debate gains in votes made by the 'for' group relative to the 'against' group. A positive ∆V means that the 'for' group gained more votes than the 'against' group, while a negative ∆V indicates that the 'against' group won more votes relative to the 'against' group.

***

# Discussion

In addition to basic data preparation, this markdown explored how a series of targeted variables (drawn from work in linguistics, psychology, and political science) influence persuasion. We selected four linguistic behaviors and personal traits for each debater -- attractiveness, hedging, negativity, and pronouns -- and explored how these factors might influence persuasiveness during a live debate. Contrary to our expectations, we found only an effect of attractiveness on persuasion, in which greater attractiveness ratings were associated with higher "for"-group persuasiveness, *even if* the attractive debater were a member of the "against" group.

After investigating these targeted effects, our next R markdown leverages much broader linguistic measures in a series of exploratory models to better understand persuasiveness.
